---
title: 'Note to self: column selection and row filtering in dplyr'
date: 2026-07-31
categories:
   - notes to self
   - dplyr
   - R
format:
  html:
    df-print: kable
---

```{r include=FALSE}
library(dplyr)
library(magrittr)
library(stringr)
```

As I'm no longer quite the spring chicken I once was, and information seems to
be harder to get to stick in the brain, here's the first in a series of small 
notes to myself (and anyone else for whom they might be useful) of things that 
I repeatedly have to look up.

If I have a data frame in R:

```{r echo=FALSE}
df <- data.frame(
  X1 = c("A1", "A2", "B1", "B2"),
  X2 = c("C1", "C2", "D1", "D2"),
  Y1 = c("E1", "E2", "F1", "F2")
)

df
```

and I want to select columns whose name begins with a string, the syntax is:

```{r}
df %>% dplyr::select(dplyr::starts_with("X"))
```

But if I want to filter to include only _rows_ whose value (in a particular column)
starts with a certain string, then I can use the `stringr` package: 

```{r}
df %>% dplyr::filter(stringr::str_starts(X1, "A"))
```